---
title: "Week 1&2 Lab Handout"
author: "Yuanchao Zhang"
date: "1/3/2018"
output: 
  html_document:
    theme: united
    toc: true
    number_sections: true
    toc_depth: 3
---

```{r setup, include=FALSE}

```

This handout covers the topics that we will go through during the lab session. The content is similar to the lecture note but with more focus on operational details. 

There are two main sections in this handout, required and optional. 

* The required section covers the topics listed in the syllabus. 
* The optional section covers some other useful topics in case the lab session ends too early. 

# Required

## Install R programming language

1. Go to <https://www.r-project.org/>. 
2. Click `download R`. 

![Command Line with Some Examples]()


```{r cars}
if (require('ggplot2')) print(123)
summary(cars)
```

# Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
